Markdown to estimate the abundance of brain cell types with the reprocessed counts from ROSMAP cohort (with TMM-VOOM normalisation)

### 2017-02-02 code/Rmd/ROSMAP_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd
Added markdown to deconvolute cell typr fraction in reprocessed RNASeq from ROSMAP

---
title: "Covariate analysis of ROSMAP reprocessed counts (with TMM and SVA normalisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

@@

knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_TMM_SVA.Rmd",
                                 parentId = "syn5570291",
                                 entityName = 'ROSMAP Reprocessed RNASeq TMM SVA (Based on cogdx)')
```

@@

library(ggplot2)
library(reshape2)
library(limma)
library(Biobase)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)

@@

@@

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570291';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of ROSMAP reprocessed RNASeq data';

thisFileName <- 'ROSMAP_RNASeq_Reprocessed_TMM_SVA.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_ID <- 'syn7750270';
ALL_USED_IDs <- COUNT_ID
COUNT_OBJ <- synGet(COUNT_ID)
COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F)

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering, cache=TRUE}
#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv"
FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx")
ContCovariates <- c("RINcontinuous","age_first_ad_dx","age_death","age_at_visit_max","pmi","educ")

@@

# Add in RIN^2 values
COVARIATES$RINcontinuous2 = COVARIATES$RINcontinuous^2
ContCovariates = c(ContCovariates, 'RINcontinuous2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.5, row.width=0.1)
```

```{r dataExplore, fig.width = 10, fig.height = 10}
# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = cogdx, y = RINcontinuous)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')

# AgeAtDeath
p[[2]] = ggplot(COVARIATES, aes(x = cogdx, y = age_death)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + theme(legend.position = 'top')

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = cogdx, y = pmi)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI') + theme(legend.position = 'top')

# Education
p[[4]] = ggplot(COVARIATES, aes(x = cogdx, y = educ)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Education') + theme(legend.position = 'top')

multiplot(plotlist = p, cols = 2)

# Batch
vcd::mosaic(~ Batch + cogdx, data = COVARIATES)

# Sex
vcd::mosaic(~ msex + cogdx, data = COVARIATES)
```
### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

@@

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(VOOM.GENE_EXPRESSION$E))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Batch, shape=cogdx, size=RINcontinuous))
p <- p + theme_bw() + theme(legend.position="right")
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Batch', 'msex', 'cogdx')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Batch', 'msex', 'cogdx')]))
```

```{r temp, include = F}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "cogdx" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on cognitive scores)

Adding Batch and msex a priori to variable selection
```{r customNorm, results='asis'}
# Primary variable of interest
postAdjustCovars = c('cogdx', 'Batch', 'msex');

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, postAdjustCovars)]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
                                                                          c(postAdjustCovars))]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### SVA 
Estimate surrogate variables using SVA package
```{r sva}
MODEL0 = DM1$design[,-grep('cogdx',colnames(DM1$design))] # Get null model by removing variable of interest
MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
MODEL1 = DM1$design
MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]

## Calculate the number of surrogate variables
NUM.SV = sva::num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1,
                     method = 'be', seed = 123456, B = 30)

SURR.VAR = sva::sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0,
                    n.sv = NUM.SV, B = 30)$sv

colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
rownames(SURR.VAR) = rownames(DM1$design)

DM1$design = cbind(MODEL1, SURR.VAR)

# Estimate voom weights
VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)

# Fit linear model using new weights and new design
VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
```

### Residual calculation
Calculate weighted residuals and add back "cogdx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("cogdx", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2)) 
p <- p + geom_point(aes(color=Batch, shape=cogdx, size=RINcontinuous))
p <- p + theme_bw() + theme(legend.position="right")
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)
p
```

```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Batch', 'msex', 'cogdx')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

```{r temp1, include=F}
dev.off()
gc()
```

```{r get.gene.symbols, cache=TRUE, eval=T}
backgroundGenes = data.frame(gene.id = rownames(COUNT)) %>%
  dplyr::mutate(ensembl_gene_id = gene.id) %>%
  tidyr::separate(ensembl_gene_id, c('ensembl_gene_id', 'position'), sep = '\\.')

@@

# Add gene names
RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('gene.id') %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('gene.id') %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```

Interpretation of cogdx scores

1. NCI, No cognitive impairment (No impaired domains)
2. MCI, Mild cognitive impairment (One impaired domain) and NO other cause of CI
3. MCI, Mild cognitive impairment (One impaired domain) AND another cause of CI
4. AD, Alzheimer's disease and NO other cause of CI (NINCDS PROB AD)
5. AD, Alzheimer's disease AND another cause of CI (NINCDS POSS AD)
6. Other dementia. Other primary cause of dementia

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=8, fig.width=16}
# Fit contrast
contrast = makeContrasts(contrasts=c("cogdx1-cogdx2",
                                     "cogdx1-cogdx4",
                                     "cogdx2-cogdx4"),
                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:3, function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC){
  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T) %>%
    rownameToFirstColumn('gene.id') %>%
    left_join(backgroundGenes) %>%
    left_join(Ensemble2HGNC) %>%
    dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, logFC, CI.L, CI.R, AveExpr, t, P.Value, adj.P.Val, B)
}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC) 
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Study = 'ROSMAP',
                Region = 'DLPFC',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')))

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
  group_by(Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05)

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction) %>%
  group_by(Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2)

kable(full_join(tmp %>%
                  dplyr::rename(DOWN_FDR_0_05 = DOWN, UP_FDR_0_05 = UP),
                tmp1 %>%
                  dplyr::rename(DOWN_FDR_0_05_FC_1.2 = DOWN, UP_FDR_0_05_FC_1.2 = UP)))

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
p = p + facet_grid(.~Comparison, scales = 'fixed')
p
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "ROSMAP Reprocessed RNASeq TMM SVA (Based on cogdx)", parentId = parentId)
annotations(CODE) = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay	 = 'Reprocessed RNAseq',

  tissueTypeAbrv	= 'DLPFC', 
  study = 'ROSMAP', 
  center = 'BROAD-RUSH',

  organism = 'HomoSapiens',
  consortium	= 'AMP-AD',

  normalizationStatus	= TRUE,
  modelSystem	= FALSE,

  normalizationType	= 'TMM and SVA'
)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'Sampleid')
write.table(COVARIATES, file = 'ROSMAP_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('ROSMAP_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix %>%
  as.matrix() %>% rownameToFirstColumn('gene.id') %>%
  write.table(file = 'ROSMAP_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('ROSMAP_DLPFC_Counts.tsv', name = 'Raw counts (filtered)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                     executed = thisFile, activityDescription = activityDescription)

# Store logCPM
write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'ROSMAP_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('ROSMAP_DLPFC_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)

# Store design matrix
write.table(DM1$design, file = 'ROSMAP_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('ROSMAP_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)

# Store residual gene expression
write.table(RESIDUAL.GENE_EXPRESSION, file = 'ROSMAP_DLPFC_Expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('ROSMAP_DLPFC_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store differential expression results
write.table(DE, file = 'ROSMAP_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('ROSMAP_DLPFC_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
```
|  *Results*                |  *SynapseID*                     |
|  -------                  |   ---------                      |
|  Covariates               |  `r COV_OBJ$properties$id`       |
|  Raw counts (log CPM)     |  `r COUNT_OBJ$properties$id`     |
|  Design Matrix            |  `r DM_OBJ$properties$id`        |
|  Residual Expression      |  `r EXP_OBJ$properties$id`       |
|  Differential Expression  |  `r DEXP_OBJ$properties$id`      |

>>>>

---
title: "Markdown to estimate the abundance of brain cell types with the reprocessed counts from ROSMAP cohort (with TMM-VOOM normalisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

@@

knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd",
                                 parentId = "syn5570291",
                                 entityName = 'ROSMAP Cell Type Abundance with Reprocessed RNASeq')
```

@@

library(ggplot2)
library(reshape2)
library(limma)
library(Biobase)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)
library(xlsx)

@@

library(parallel)
library(doParallel)
library(doMC)
library(foreach)

cl = makeCluster(detectCores() - 2)
registerDoParallel(cl)

@@

# Source external files
source('../R/CIBERSORT.R')

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5570291';
activityName = 'Cell type abundance in ROSMAP';
activityDescription = 'Cell type abundance in ROSMAP with TMM-VOOM normalised reprocessed RNASeq counts';

thisFileName <- 'ROSMAP_RNASeq_Reprocessed_TMM_VOOM_CellProp.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='celltype')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_ID <- 'syn7750270';
ALL_USED_IDs <- COUNT_ID
COUNT_OBJ <- synGet(COUNT_ID)
COUNT <- read.table(COUNT_OBJ@filePath,header=T,sep='\t',check.names = F)

@@

### Get Covariates
```{r covariates.clustering, cache=TRUE}
#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv"
FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx")
ContCovariates <- c("RINcontinuous","age_first_ad_dx","age_death","age_at_visit_max","pmi","educ")

@@

# Add in RIN^2 values
COVARIATES$RINcontinuous2 = COVARIATES$RINcontinuous^2
ContCovariates = c(ContCovariates, 'RINcontinuous2')
```

### CPM Normalisation
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

@@

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=NULL, plot=T)
```

### Get cell type specific signatures and expression data from synapse
```{r get.signatures}
ALL_USED_IDs = c(ALL_USED_IDs, 'syn8068511', 'syn8068475')
# Signatures from Zhang,Y et al., 2014
sig.zhang = downloadFile('syn8068511') %>%
  dlply(.(Set), .fun = function(x){return(unique(x$Gene))})

# Signatures from Darmanis, S et al., 2015
sig.dar.obj = synGet('syn8068475')  
sig.dar.unbiased = read.xlsx(sig.dar.obj@filePath, sheetName = 'Unbiased') %>%
  lapply(function(x){as.character(unique(x))})
sig.dar.unbiased = sig.dar.unbiased[1:10]

# Extract all signatures
sig = list()
sig$OPC = unique(c(sig.zhang$OPC, sig.dar.unbiased$Cluster.1))
sig$Oligodendrocyte = unique(c(sig.zhang$MyelinOligos, sig.zhang$NewOligos, sig.dar.unbiased$Cluster.3))
sig$Neuron = unique(c(sig.zhang$Neuron, sig.dar.unbiased$Cluster.6, sig.dar.unbiased$Cluster.7,
                      sig.dar.unbiased$Cluster.10, sig.dar.unbiased$Cluster.2))
sig$Endothelial = unique(c(sig.zhang$Endothelial, sig.dar.unbiased$Cluster.8))
sig$Astrocyte = unique(c(sig.zhang$Astrocyte, sig.dar.unbiased$Cluster.4))
sig$Microglia = unique(c(sig.zhang$Astrocyte, sig.dar.unbiased$Cluster.5))

sig = ldply(sig, .fun = function(x){as.data.frame(x)})
colnames(sig) = c('CellType','Gene.Name')

# Get reference expression data from Darmanis., S et al., 2015
ALL_USED_IDs = c(ALL_USED_IDs, 'syn8077142', 'syn8077191')
ref.expr = downloadFile('syn8077142')
rownames(ref.expr) = ref.expr$hgnc_symbol
ref.expr$hgnc_symbol = NULL

ref.cellTypes = downloadFile('syn8077191') %>%
  dplyr::select(SampleID, CellType)

ref.expr1 = list()
for(celltype in unique(ref.cellTypes$CellType)){
  ids = ref.cellTypes$SampleID[ref.cellTypes$CellType == celltype]
  ref.expr1[[celltype]] = log2(rowSums(2^ref.expr[,ids], na.rm = T)) %>%
    rownameToFirstColumn('hgnc_symbol') %>%
    plyr::rename(c('DF' = celltype))
}
ref.expr1 = join_all(ref.expr1)
```

```{r get.gene.symbols, cache=TRUE, eval=T}
backgroundGenes = data.frame(gene.id = rownames(VOOM.GENE_EXPRESSION$E)) %>%
  dplyr::mutate(ensembl_gene_id = gene.id) %>%
  tidyr::separate(ensembl_gene_id, c('ensembl_gene_id', 'position'), sep = '\\.')

@@

# Add gene names
EXPR <- VOOM.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('gene.id') %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  filter(!duplicated(gene.id))
rownames(EXPR) = EXPR$gene.id

EXPR1 = WGCNA::collapseRows(EXPR %>% dplyr::select(-ensembl_gene_id, -position, -hgnc_symbol, -gene.id), 
                            EXPR$hgnc_symbol, 
                            EXPR$gene.id)$datETcollapsed

ind = intersect(rownames(EXPR1), ref.expr1$hgnc_symbol)
EXPR1 = EXPR1 %>% 
  rownameToFirstColumn('hgnc_symbol') %>%
  filter(hgnc_symbol %in% ind)
ref.expr1 = ref.expr1 %>%
  filter(hgnc_symbol %in% ind)

write.table(ref.expr1, file = 'ref.expr.txt', sep = '\t', quote=F, row.names = F)
write.table(EXPR1[,1:20], file = 'mixture.expr.txt', sep = '\t', quote=F, row.names = F)

cellprop = CIBERSORT('ref.expr.txt', 'mixture.expr.txt', perm = 100) # Obtain CIBERSORT.R function from http://cibersort.stanford.edu/
stopCluster(cl)
```

```{r plot.cellprop}
cellprop1 = as.data.frame(cellprop)
cellprop1$cogdx = COVARIATES[rownames(cellprop), 'cogdx']
cellprop1 = cellprop1 %>% 
  rownameToFirstColumn('SampleID') %>%
  dplyr::select(-`P-value`,-Correlation, -RMSE) %>%
  tidyr::gather(Celltype, Fraction, -SampleID, -cogdx) %>%
  dplyr::arrange(desc(cogdx))

p = ggplot(cellprop1, aes(x = SampleID, y = Fraction, fill = Celltype, color = cogdx)) + geom_bar(stat = 'identity')
p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
# p = p + facet_grid(.~cogdx)
p

p = ggplot(cellprop1, aes(x = Celltype, y = Fraction, color = cogdx)) + geom_boxplot()
p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p
```
### Synapse store
```{r syn.store, cache=TRUE}
CODE = Folder(name = 'ROSMAP Cell Type Abundance with Reprocessed RNASeq', parentId = parentId)
CODE = synStore(CODE)

cellprop %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'ROSMAP_CellFraction_Reproc.tsv', row.names = F, sep = '\t', quote = F)
obj = File('ROSMAP_CellFraction_Reproc.tsv', name = 'ROSMAP Cell Fraction', parentId = CODE$properties$id)
obj = synStore(obj, activityName = activityName, activityDescription = activityDescription,
               used = ALL_USED_IDs, executed = thisFile)
```
